---
title: "RMS Titanic"
author: "Haley Pelletier"
date: 2020-07-16
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(viridis)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
glimpse(df_titanic)
summary(df_titanic)
```

**Observations**:

Variables:

- Class (char)
- Sex (char)
- Age (char)
- Survived (char)
- n (double)

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
# Count by crew status
total_passengers <-
  df_titanic %>%
    summarize(total = sum(n))

crew_passengers <-
  df_titanic %>%
    filter(Class == "Crew") %>%
    summarize(total = sum(n))

non_crew_passengers <- total_passengers - crew_passengers

total_passengers
crew_passengers
non_crew_passengers

# Count by class
first_class_passengers <-
  df_titanic %>%
    filter(Class == "1st") %>%
    summarize(total = sum(n))

second_class_passengers <-
  df_titanic %>%
    filter(Class == "2nd") %>%
    summarize(total = sum(n))

third_class_passengers <-
  df_titanic %>%
    filter(Class == "3rd") %>%
    summarize(total = sum(n))

first_class_passengers
second_class_passengers
third_class_passengers

# Count by sex
male_non_crew_passengers <-
  df_titanic %>%
    filter(Sex == "Male", Class != "Crew") %>%
    summarize(total = sum(n))

female_non_crew_passengers <- non_crew_passengers - male_non_crew_passengers

male_non_crew_passengers
female_non_crew_passengers

# Count by age
child_passengers <-
  df_titanic %>%
    filter(Age == "Child") %>%
    summarize(total = sum(n))

third_class_child_passengers <-
  df_titanic %>%
    filter(Age == "Child", Class == "3rd") %>%
    summarize(total = sum(n))

child_passengers
third_class_child_passengers
```

**Observations**:

- From the Wikipedia page: 
  - "Titanic had around **885** crew members on board for her maiden voyage."
  - "Titanic's passengers numbered approximately **1,317** people: **324** in First Class, **284** in Second Class, and **709** in Third Class. Of these, **869** (66%) were male and **447** (34%) female. There were **107** children aboard, the largest number of whom were in Third Class."
- The total number of passengers in the dataset (**2201**) is just one less than the total according to Wikipedia (885 crew + 1317 non-crew = **2202**).
  - The number of crew members in the dataset (**885**) is the same as Wikipedia, so that means there is one non-crew passenger missing from the dataset.
- The number of First Class (**325**), Second Class (**285**), and Third Class (**706**) passengers in the dataset is only slightly different from Wikipedia.
  - The difference includes the one missing passenger we already know about and suggests that two additional passengers are miscategorized.
- The number of male (**869**) and female (**447**) non-crew passengers in the dataset is exactly the same as on Wikipedia.
- The number of children (**109**) in the dataset is slightly different from Wikipedia.
  - This suggests that two passengers are miscategorized as children, or perhaps the age cut-off is different for each source.
  - Wikipedia also says that the largest number of children were in Third Class. That holds true in the dataset (79/109 = **72.5%**).


__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
# Total
df_titanic %>%
  ggplot() +
    geom_col(aes(x = Class, y = n, fill = Sex), position = "dodge") +
    ggtitle("Total Passengers") +
    theme(plot.title = element_text(hjust = 0.5))

# Survived only
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
    geom_col(aes(x = Class, y = n, fill = Sex), position = "dodge") +
    ggtitle("Surviving Passengers") +
    theme(plot.title = element_text(hjust = 0.5))
```

**Observations**:

- Many more male crew members survived than female crew members. 
  - This is because there were many more male crew members in total.
- In First and Second class, many more women survived than men.
  - This can't be explained by total number, because there were more men than women in these classes.
  - Likely, this can be explained by this tidbit from Wikipedia: "A disproportionate number of men were left aboard because of a 'women and children first' protocol for loading lifeboats."
- In Third class, the numbers are about the same.
  - From Wikipedia, regarding the lifeboats: "Third-class passengers were largely left to fend for themselves, causing many of them to become trapped below decks as the ship filled with water."
  - Because the Third Class passengers were not following the "women and children first" lifeboat protocol, the survivors were not disproportionately women.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task}
df_prop %>%
  ggplot() +
    geom_col(aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
    ggtitle("Total Passengers") +
    theme(plot.title = element_text(hjust = 0.5))

df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
    geom_col(aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
    ggtitle("Surviving Passengers") +
    theme(plot.title = element_text(hjust = 0.5))
```

**Observations**:

- These plots don't really make sense as-is? I can't fgure out exactly why, but I think they need to be broken out by Age in additon to Class and Sex.

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
    geom_col(aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
    facet_grid(cols = vars(Age)) +
    ggtitle("Surviving Passengers") +
    theme(plot.title = element_text(hjust = 0.5))
```

**Observations**:

- This makes more sense than the plots for q4! 
- Here, we can clearly see that in the First and Second Classes (and among the Crew), all of the children and the majority of the women survived. 
- We can also see that in Third Class, where they did not have the same access to lifeboats, significantly smaller proportions of women and children survived.


__Bonus__

```{r bonus}
df_prop %>%
  filter(Class != "Crew", Survived == "Yes") %>%
  ggplot() +
    geom_point(mapping = aes(x = Age, y = Sex, color = Prop), size = 30) +
    scale_color_viridis(option = "cividis") +
    ggtitle("(Non-Crew) Women and Children More Likely to Survive than Men") +
    theme(plot.title = element_text(hjust = 0.5))
```

**Observations**:

- Trying out another visualization to show that, across the passenger classes, more women and children survived than men.


# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
